Conditional selectivity for triplestores

### 2012-09-06 conditional/trunk/selectivity-basic.tex
conditional triples in a pattern
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@363 43ef1190-90fd-4dd7-a769-d0121f02b80e
first derived selectivity
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@365 43ef1190-90fd-4dd7-a769-d0121f02b80e

It is easy to point out cases where the independence assumption is
unlikely to hold: If a predicate is \rdfterm{rdf:type}, then only
classes are likely objects, and their frequency  would likely be
different from e.g. literals.

@@

In section~\ref{sec:cond}, we saw

>>>>

It is easy to point out cases where the independence assumption is
unlikely to hold: If a predicate is \rdfterm{rdf:type}, then only
classes are likely objects, and their frequency would likely be
different from e.g. literals. Moreover, it is easy to see when triples
in a basic graph pattern are likely to be dependent: If one triple has
\rdfterm{rdf:type} as predicate and \rdfterm{foaf:Person} as object,
the next triple is much more likely to contain the very
common predicate \rdfterm{foaf:name} than the esoteric
\rdfterm{foaf:dnaChecksum}.

@@

In section~\ref{sec:cond}, we derived an equation for conditional
selectivity, i.e. RDF terms are not independent of each
other.

In the follwing, we denote a bounded term with a lowercase letter, $s$
for a subject, $p$ for a predicate and $o$ for object, and variables
are denoted by a question mark, $?$. For example, $p =
\rdfterm{rdf:type}$. We also adopt SPLENDIDs convention of denoting
the number of all triples in certain data source $|d|$.

We shall now systematically go through different triple patterns and
see how the selectivity can be derived from VoID descriptions per the
most recent vocabulary\cite{voidnote}:

\paragraph{Subject unbound, object and predicate bound}

In this case, we can use Eq.~\ref{eq:sel} directly. We again note that
since the the unbound subject will always have selectivity of
unity.
Eq.~\ref{eq:sel} thus reduces to:

\begin{equation}\label{eq:selsubjun}
sel(T_{?~p~o}) = sel(O | P) \cdot sel(P) . 
\end{equation}

To relate this to VoID, we note that VoID includes the concept of
\emph{property partitions} (see section~4.5 in \cite{voidnote}): ``A
property-based partition contains only those triples of a dataset that
use a particular predicate.'', and since a partition is itself a
dataset, the same statistical properties can be applied, such as
\rdfterm{void:triples}, \rdfterm{void:distinctSubjects} and
\rdfterm{void:distinctObjects}.

Since property partitions detail statistics given a certain predicate,
they can be used in the conditional case. We can therefore write:

\begin{equation}\label{eq:selsubjunp}
sel(P = p) = \frac{\rdfterm{void:propertyPartition/void:triples}}{|d|}
\end{equation}
and
\begin{equation}\label{eq:selsubjunop}
sel(O = o|P = p) =
\frac{\rdfterm{void:propertyPartition/void:distinctObjects}}{\rdfterm{void:propertyPartition/void:triples}} .
\end{equation}

Thus,
\begin{equation}\label{eq:selsubjunop}
sel(T_{?~p~o}) =
\frac{\rdfterm{void:propertyPartition/void:distinctObjects}}{|d|} .
\end{equation}\todo{wtf}
